Evaluating Perceptual Judgements on 3D Printed Bar Charts

### 2024-03-21 jds.rmd
Updated paper based on reviewer feedback

res_all = results %>% 
  left_join(trueRatios, by = 'fileID') %>% 
  mutate(response = log2(abs(byHowMuch - Height*100) + 1/8),
         subject = paste0(nickname, participantUnique),
         ratioLabel = round(100*Height, 1)) %>% 
  filter(subject %in% valid.users$subject) %>% 
  arrange(appStartTime)

res <- res_all %>% 
  filter(whichIsSmaller == 'Triangle (▲)',
         as.Date.POSIXct(appStartTime) <= '2023-05-22') %>% 
  #Only verified results from before the start of SDSS 
  filter(!is.na(ratio))

@@

The accuracy of graphical forms has been studied for almost a century [@vonhuhnFurtherStudiesGraphic1927;@eellsRelativeMeritsCircles1926;@croxtonGraphicComparisonsBars1932;@croxtonBarChartsCircle1927], as new ways of representing information evolve, we must revisit old studies to determine whether these representations have the same limitations as previous versions. This is particularly true in areas like graphics which are affected by the immense technological innovation in hardware and software which has taken place since the early 1990s.

@@

There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data [@huffHowLieStatistics1954]. 
However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since Heer's replication study [-@heerCrowdsourcingGraphicalPerception2010b] in 2010.  
Computer graphics and rendering technology have developed quickly, along with the hardware necessary to support these software developments. 
As a result, we have much more natural virtual rendering of 3D objects, and we can also print graphics in three dimensions using mass-market, relatively inexpensive 3D printers, moving artificial charts into a more natural, physical setting.
As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because 3D printed charts provide the opportunity to make data graphics accessible to those with limited or absent vision [@fleetHowTactileGraphics2023].

@@

A diagram of the experimental design is provided in \Cref{fig:studyDesign}. The study is set up as a modified randomized incomplete block design, comprised of 3 blocks (presentation mode), 7 bar ratios (5 are selected for each participant). At each combination of ratio and presentation type, the comparison type, adjacent (Type 1) or separated (Type 3), is randomly selected. This process was used to create 21 kits of 5 3D printed charts and 10 virtual charts. There is thus a slight deviation from a completely randomized incomplete block design, in that all possible combinations of ratios, presentation modes, and comparison types are not present in our design. However, this approach provided the best experimental design given the constraints of managing physical stimuli; it took nearly a month of continuous 3D printing to create all of the charts for the kits that we assembled. A dataset containing the kit composition breakdown is available in the github repository for this project.

@@

A total of 48 participants completed the study; demographics are shown in \Cref{fig:demographics}.

@@

Each graph (or prompt, in the case of 3D printed charts) in the applet had two corresponding questions for participants to answer: first, participants were to identify the smaller bar by shape, and then, participants were to estimate the ratio of the size of the smaller bar to the size of the larger bar, as shown in \Cref{fig:experiment3dRender}. 
It should be noted that we chose to use a slider for entering ratio estimates - in part, this was intended to alleviate rounding effects which occur when participants enter numbers on their own or get numeric feedback from slider entries [@ruudUncertaintyCausesRounding2014a;@maineriSliderBarsMultiDevice2021]. 
Another benefit of this modification is that it should reduce participants' cognitive load, allowing them to focus on the ratio between bars rather than the numerical translation process.
This differs from @clevelandGraphical1984, but maintains the spirit of the study while using methods which were not available for data entry at the time.

@@

All responses ($n = `r nrow(res_all) - nrow(res)`$) that incorrectly identified the smaller bar were removed from the study before analysis; this serves as a basic attention check. In addition, one trial was removed because the participant did not enter a valid ID for a 3D printed chart.

@@

A generalized additive model with random effects for subject was fit using the `mgcv` package [@mgcv1;@mgcv2;@mgcv3]. Spline terms by comparison type and display type were included for the ratio variable, to account for indications of nonlinearity from \Cref{fig:midmeans-log-errors}; cumulative predictions for fixed effects are shown in \Cref{fig:gam-pred-chart}. It is clear that there are differences between the errors for separated comparisons and adjacent comparisons; adjacent comparisons seem to have roughly similar errors for all ratio values, while separated comparisons seem to have lower errors for points near 0, .5, and 1, with higher errors as the distance increases from one of these anchor points; this effect has been explored in @hollandsBiasProportionJudgments2000. 
Residuals from the model indicate that there is some left skew in the tails, but this is not extreme; code for this analysis can be found in the supplemental materials on github.

@@

While we were initially concerned that this effect was due to the slider initial setting of 0.5, we did examine this hypothesis; only `r sum(res$byHowMuch == 50)` trials (`r sprintf("%.2f%%", mean(res$byHowMuch==50)*100)`) had slider values at exactly 0.5, and all but `r sum(res$byHowMuch == 50 & !res$ratioLabel %in% c(46.4, 56.2))` were for ratios on either side of 0.5 (.464, .562). This suggests that while a few participants did take the strategy of using the default value for the slider, they did so strategically, not moving the slider only when the ratio was perceived to be sufficiently close to 0.5. In addition, trials where participants utilized this strategy were evenly split between separated and adjacent comparisons, suggesting that this anchoring process is not the reason for the difference in spline fits for the different comparison types. It seems more likely that this effect is at least partially due instead to lower errors around the extremes of the ratio spectrum. A follow-up experiment with more ratios and in particular ratio values closer to suspected anchor points  could explore this relationship in more depth - with only 7 different ratios, it is difficult to do more than note the appearance of an effect and speculate about the anchor points, particularly given that we have used spline models with 5 knots to fit the smooth.

@@

The first potential explanation is that this study is underpowered - the effect size is small, and our 48 participants were insufficient; the original study included 51 participants, which is slightly larger. However, we should note that the analysis methods used in @heerCrowdsourcingGraphicalPerception2010b and @clevelandGraphical1984 do not provide numerical tests of statistical significance, instead defaulting to graphical displays which include confidence intervals.

>>>>

res_all = results %>% 
  left_join(trueRatios, by = 'fileID') %>% 
  mutate(response = log2(abs(byHowMuch - Height*100) + 1/8),
         subject = paste0(nickname, participantUnique),
         ratioLabel = round(100*Height, 1)) %>% 
  filter(subject %in% valid.users$subject
         , as.Date.POSIXct(appStartTime) <= '2023-05-22'
         ) %>% 
  arrange(appStartTime)

valid.users <- valid.users %>% 
  filter(subject %in% res_all$subject)

res <- res_all %>% 
  filter(whichIsSmaller == 'Triangle (▲)'
         , as.Date.POSIXct(appStartTime) <= '2023-05-22'
         ) %>% 
  #Only verified results from before the start of SDSS 
  filter(!is.na(ratio))

@@

The accuracy of graphical forms has been studied for almost a century [@vonhuhnFurtherStudiesGraphic1927;@eellsRelativeMeritsCircles1926;@croxtonGraphicComparisonsBars1932;@croxtonBarChartsCircle1927]; as new ways of representing information evolve, we must revisit old studies to determine whether these representations have the same limitations as previous versions. This is particularly true in areas like graphics which are affected by the immense technological innovation in hardware and software which has taken place since the early 1990s.

@@

There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data [@huffHowLieStatistics1954]. 
However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since Heer's replication study [-@heerCrowdsourcingGraphicalPerception2010b] in 2010.  
Computer graphics and rendering technology have developed quickly, along with the hardware necessary to support these software developments. 
As a result, we have much more natural virtual rendering of 3D objects, and we can also print graphics in three dimensions using mass-market, relatively inexpensive 3D printers, moving artificial charts into a more natural, physical setting.
As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because 3D printed charts could eventually provide the opportunity to make data graphics accessible to those with limited or absent vision [@fleetHowTactileGraphics2023].

@@

A diagram of the experimental design is provided in \Cref{fig:studyDesign}. The study is set up as a modified randomized incomplete block design, comprised of 3 blocks (presentation mode) and 7 bar ratios (5 are selected for each participant). At each combination of ratio and presentation type, the comparison type, adjacent (Type 1) or separated (Type 3), is randomly selected. This process was used to create 21 kits of 5 3D printed charts and 10 virtual charts. There is thus a slight deviation from a completely randomized incomplete block design, in that all possible combinations of ratios, presentation modes, and comparison types are not present in our design. However, this approach provided the best experimental design given the constraints of managing physical stimuli; it took nearly a month of continuous 3D printing to create all of the charts for the kits that we assembled. A dataset containing the kit composition breakdown is available in the github repository for this project.

@@

A total of `r nrow(valid.users)` participants completed the study; demographics are shown in \Cref{fig:demographics}.
The participants provided a total of `nrow(res_all)` responses.

@@

Each graph (or prompt, in the case of 3D printed charts) in the applet had two corresponding questions for participants to answer: first, participants were to identify the smaller bar by shape, and then, participants were to estimate the ratio of the size of the smaller bar to the size of the larger bar, as shown in \Cref{fig:experiment3dRender}. 
It should be noted that we chose to use a slider without numeric identifiers or tick marks for entering ratio estimates - in part, this was intended to alleviate rounding effects which occur when participants enter numbers on their own or get numeric feedback from slider entries [@ruudUncertaintyCausesRounding2014a;@maineriSliderBarsMultiDevice2021]. 
Another benefit of this modification is that it should reduce participants' cognitive load, allowing them to focus on the ratio between bars rather than the numerical translation process.
This differs from @clevelandGraphical1984, but maintains the spirit of the study while using methods which were not available for data entry at the time.

@@

All responses that incorrectly identified the smaller bar ($n = `r nrow(res_all) - nrow(res)`$) were removed from the study before analysis; this serves as a basic attention check and leaves `r nrow(res)` valid responses. In addition, one response was removed because the participant did not enter a valid ID for a 3D printed chart.

@@

A generalized additive model with random effects for subject was fit using the `mgcv` package [@mgcv1;@mgcv2;@mgcv3]. Spline terms by comparison type and display type were included for the ratio variable, to account for indications of nonlinearity from \Cref{fig:midmeans-log-errors}; cumulative predictions for fixed effects are shown in \Cref{fig:gam-pred-chart}. It is clear that there are differences between the errors for separated comparisons and adjacent comparisons; adjacent comparisons seem to have roughly similar errors for all ratio values, while separated comparisons seem to have lower errors for points near 0, .5, and 1, with higher errors as the distance increases from one of these anchor points; this effect has been explored in @hollandsBiasProportionJudgments2000. 
Residuals from the model indicate that there is some left skew in the tails, but this is not extreme; code for this analysis can be found in the supplemental materials on GitHub.

@@

While we were initially concerned that this effect was due to the slider initial setting of 0.5, we did examine this hypothesis; only `r sum(res$byHowMuch == 50)` responses (`r sprintf("%.2f%%", mean(res$byHowMuch==50)*100)`) had slider values at exactly 0.5, and all but `r sum(res$byHowMuch == 50 & !res$ratioLabel %in% c(46.4, 56.2))` were for ratios on either side of 0.5 (.464, .562). This suggests that while a few participants did take the strategy of using the default value for the slider, they did so strategically, not moving the slider only when the ratio was perceived to be sufficiently close to 0.5. In addition, trials where participants utilized this strategy were evenly split between separated and adjacent comparisons, suggesting that this anchoring process is not the reason for the difference in spline fits for the different comparison types. It seems more likely that this effect is at least partially due instead to lower errors around the extremes of the ratio spectrum. A follow-up experiment with more ratios and in particular ratio values closer to suspected anchor points  could explore this relationship in more depth - with only 7 different ratios, it is difficult to do more than note the appearance of an effect and speculate about the anchor points, particularly given that we have used spline models with 5 knots to fit the smooth.

@@

The first potential explanation is that this study is underpowered - the effect size is small, and our `r nrow(valid.users)` participants were insufficient; the original study included 51 participants, which is slightly larger. However, we should note that the analysis methods used in @heerCrowdsourcingGraphicalPerception2010b and @clevelandGraphical1984 do not provide numerical tests of statistical significance, instead defaulting to graphical displays which include confidence intervals.

### 2024-03-21 jds.tex
Updated paper based on reviewer feedback

The accuracy of graphical forms has been studied for almost a century \citep{vonhuhnFurtherStudiesGraphic1927, eellsRelativeMeritsCircles1926, croxtonGraphicComparisonsBars1932, croxtonBarChartsCircle1927}, as new ways of representing information evolve, we must revisit old studies to determine whether these representations have the same limitations as previous versions. This is particularly true in areas like graphics which are affected by the immense technological innovation in hardware and software which has taken place since the early 1990s.

@@

There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data \citep{huffHowLieStatistics1954}.
However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since Heer's replication study \citeyearpar{heerCrowdsourcingGraphicalPerception2010b} in 2010.\\
Computer graphics and rendering technology have developed quickly, along with the hardware necessary to support these software developments.
As a result, we have much more natural virtual rendering of 3D objects, and we can also print graphics in three dimensions using mass-market, relatively inexpensive 3D printers, moving artificial charts into a more natural, physical setting.
As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because 3D printed charts provide the opportunity to make data graphics accessible to those with limited or absent vision \citep{fleetHowTactileGraphics2023}.

@@

A total of 48 participants completed the study; demographics are shown in \Cref{fig:demographics}.

@@

Each graph (or prompt, in the case of 3D printed charts) in the applet had two corresponding questions for participants to answer: first, participants were to identify the smaller bar by shape, and then, participants were to estimate the ratio of the size of the smaller bar to the size of the larger bar, as shown in \Cref{fig:experiment3dRender}.
It should be noted that we chose to use a slider for entering ratio estimates - in part, this was intended to alleviate rounding effects which occur when participants enter numbers on their own or get numeric feedback from slider entries \citep{ruudUncertaintyCausesRounding2014a, maineriSliderBarsMultiDevice2021}.
Another benefit of this modification is that it should reduce participants' cognitive load, allowing them to focus on the ratio between bars rather than the numerical translation process.
This differs from \citet{clevelandGraphical1984}, but maintains the spirit of the study while using methods which were not available for data entry at the time.

@@

All responses (\(n = 33\)) that incorrectly identified the smaller bar were removed from the study before analysis; this serves as a basic attention check. In addition, one trial was removed because the participant did not enter a valid ID for a 3D printed chart.

@@

\caption{\label{tab:gam-smooth}Approximate significance of smooth terms in gam model.}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
Smooth & edf & Ref.df & F & p-value\\
\midrule
s(Ratio):plot2dDigital & 0.0005 & 0.0009 & 0.0000 & 0.5000\\
s(Ratio):plot3dDigital & 1.7934 & 2.2056 & 1.2807 & 0.2481\\
s(Ratio):plot3dPrint & 1.0003 & 1.0005 & 2.4488 & 0.1183\\
s(Ratio):typeSeparated & 4.2371 & 4.7338 & 5.8074 & 0.0001\\
s(Ratio):typeAdjacent & 1.0007 & 1.0014 & 3.5452 & 0.0604\\
\addlinespace
s(participant) & 27.0449 & 34.0000 & 4.7199 & 0.0000\\
\bottomrule
\end{tabular}
\end{table}

A generalized additive model with random effects for subject was fit using the \texttt{mgcv} package \citep{mgcv1, mgcv2, mgcv3}. Spline terms by comparison type and display type were included for the ratio variable, to account for indications of nonlinearity from \Cref{fig:midmeans-log-errors}; cumulative predictions for fixed effects are shown in \Cref{fig:gam-pred-chart}. It is clear that there are differences between the errors for separated comparisons and adjacent comparisons; adjacent comparisons seem to have roughly similar errors for all ratio values, while separated comparisons seem to have lower errors for points near 0, .5, and 1, with higher errors as the distance increases from one of these anchor points; this effect has been explored in \citet{hollandsBiasProportionJudgments2000}.
Residuals from the model indicate that there is some left skew in the tails, but this is not extreme; code for this analysis can be found in the supplemental materials on github.

@@

While we were initially concerned that this effect was due to the slider initial setting of 0.5, we did examine this hypothesis; only 42 trials (8.81\%) had slider values at exactly 0.5, and all but 6 were for ratios on either side of 0.5 (.464, .562). This suggests that while a few participants did take the strategy of using the default value for the slider, they did so strategically, not moving the slider only when the ratio was perceived to be sufficiently close to 0.5. In addition, trials where participants utilized this strategy were evenly split between separated and adjacent comparisons, suggesting that this anchoring process is not the reason for the difference in spline fits for the different comparison types. It seems more likely that this effect is at least partially due instead to lower errors around the extremes of the ratio spectrum. A follow-up experiment with more ratios and in particular ratio values closer to suspected anchor points could explore this relationship in more depth - with only 7 different ratios, it is difficult to do more than note the appearance of an effect and speculate about the anchor points, particularly given that we have used spline models with 5 knots to fit the smooth.

@@

The first potential explanation is that this study is underpowered - the effect size is small, and our 48 participants were insufficient; the original study included 51 participants, which is slightly larger. However, we should note that the analysis methods used in \citet{heerCrowdsourcingGraphicalPerception2010b} and \citet{clevelandGraphical1984} do not provide numerical tests of statistical significance, instead defaulting to graphical displays which include confidence intervals.

>>>>

The accuracy of graphical forms has been studied for almost a century \citep{vonhuhnFurtherStudiesGraphic1927, eellsRelativeMeritsCircles1926, croxtonGraphicComparisonsBars1932, croxtonBarChartsCircle1927}; as new ways of representing information evolve, we must revisit old studies to determine whether these representations have the same limitations as previous versions. This is particularly true in areas like graphics which are affected by the immense technological innovation in hardware and software which has taken place since the early 1990s.

@@

There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data \citep{huffHowLieStatistics1954}.
However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since Heer's replication study \citeyearpar{heerCrowdsourcingGraphicalPerception2010b} in 2010.\\
Computer graphics and rendering technology have developed quickly, along with the hardware necessary to support these software developments.
As a result, we have much more natural virtual rendering of 3D objects, and we can also print graphics in three dimensions using mass-market, relatively inexpensive 3D printers, moving artificial charts into a more natural, physical setting.
As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because 3D printed charts could eventually provide the opportunity to make data graphics accessible to those with limited or absent vision \citep{fleetHowTactileGraphics2023}.

@@

A total of 38 participants completed the study; demographics are shown in \Cref{fig:demographics}.
The participants provided a total of \texttt{nrow(res\_all)} responses.

@@

Each graph (or prompt, in the case of 3D printed charts) in the applet had two corresponding questions for participants to answer: first, participants were to identify the smaller bar by shape, and then, participants were to estimate the ratio of the size of the smaller bar to the size of the larger bar, as shown in \Cref{fig:experiment3dRender}.
It should be noted that we chose to use a slider without numeric identifiers or tick marks for entering ratio estimates - in part, this was intended to alleviate rounding effects which occur when participants enter numbers on their own or get numeric feedback from slider entries \citep{ruudUncertaintyCausesRounding2014a, maineriSliderBarsMultiDevice2021}.
Another benefit of this modification is that it should reduce participants' cognitive load, allowing them to focus on the ratio between bars rather than the numerical translation process.
This differs from \citet{clevelandGraphical1984}, but maintains the spirit of the study while using methods which were not available for data entry at the time.

@@

All responses that incorrectly identified the smaller bar (\(n = 18\)) were removed from the study before analysis; this serves as a basic attention check and leaves 477 valid responses. In addition, one response was removed because the participant did not enter a valid ID for a 3D printed chart.

@@

\caption{\label{tab:gam-smooth}Approximate significance of smooth terms in gam model.}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
Smooth & edf & Ref.df & F & p-value\\
\midrule
s(Ratio):plot2dDigital & 0.0005 & 0.0009 & 0.0000 & 0.5000\\
s(Ratio):plot3dDigital & 1.7934 & 2.2056 & 1.2807 & 0.2481\\
s(Ratio):plot3dPrint & 1.0003 & 1.0005 & 2.4488 & 0.1183\\
s(Ratio):typeSeparated & 4.2371 & 4.7338 & 5.8074 & 0.0001\\
s(Ratio):typeAdjacent & 1.0007 & 1.0014 & 3.5452 & 0.0604\\
\addlinespace
s(participant) & 27.0449 & 34.0000 & 4.7276 & 0.0000\\
\bottomrule
\end{tabular}
\end{table}

A generalized additive model with random effects for subject was fit using the \texttt{mgcv} package \citep{mgcv1, mgcv2, mgcv3}. Spline terms by comparison type and display type were included for the ratio variable, to account for indications of nonlinearity from \Cref{fig:midmeans-log-errors}; cumulative predictions for fixed effects are shown in \Cref{fig:gam-pred-chart}. It is clear that there are differences between the errors for separated comparisons and adjacent comparisons; adjacent comparisons seem to have roughly similar errors for all ratio values, while separated comparisons seem to have lower errors for points near 0, .5, and 1, with higher errors as the distance increases from one of these anchor points; this effect has been explored in \citet{hollandsBiasProportionJudgments2000}.
Residuals from the model indicate that there is some left skew in the tails, but this is not extreme; code for this analysis can be found in the supplemental materials on GitHub.

@@

While we were initially concerned that this effect was due to the slider initial setting of 0.5, we did examine this hypothesis; only 42 responses (8.81\%) had slider values at exactly 0.5, and all but 6 were for ratios on either side of 0.5 (.464, .562). This suggests that while a few participants did take the strategy of using the default value for the slider, they did so strategically, not moving the slider only when the ratio was perceived to be sufficiently close to 0.5. In addition, trials where participants utilized this strategy were evenly split between separated and adjacent comparisons, suggesting that this anchoring process is not the reason for the difference in spline fits for the different comparison types. It seems more likely that this effect is at least partially due instead to lower errors around the extremes of the ratio spectrum. A follow-up experiment with more ratios and in particular ratio values closer to suspected anchor points could explore this relationship in more depth - with only 7 different ratios, it is difficult to do more than note the appearance of an effect and speculate about the anchor points, particularly given that we have used spline models with 5 knots to fit the smooth.

@@

The first potential explanation is that this study is underpowered - the effect size is small, and our 38 participants were insufficient; the original study included 51 participants, which is slightly larger. However, we should note that the analysis methods used in \citet{heerCrowdsourcingGraphicalPerception2010b} and \citet{clevelandGraphical1984} do not provide numerical tests of statistical significance, instead defaulting to graphical displays which include confidence intervals.
